A meta analysis of tournaments and an evaluation of performance in the Iterated Prisoner's Dilemma.

### 2019-12-28 paper/main.tex
gets consistent with the use of Axelrod-Python library/APL/citation

Though only a select pieces of work have been discussed, the IPD literature is
rich, and new strategies and competitions are being published every year~\cite{Glynatsi2019}. The
question, however, still remains the same: what is the best way to play the
game? Compared to other works, whereas a few selected strategies are evaluated
on a small number of tournaments, this manuscript evaluates the performance of \numberofstrategies
strategies in \numberofalltournaments tournaments. These tournaments do not
consist by just standard round robin tournaments, but also by tournaments with noise
and tournaments with a probabilistic ending. The later part of the paper, evaluates
the impact of features on the performance of the strategies using modern
machine learning techniques. These features include measures regarding a
strategy's behaviour and measures regarding the tournaments. The data set used
in this work has been made publicly available~\cite{data} and can be used
for further analysis and insights.

The different tournament types as well as the data collection, which is made
possible due to an open source package called Axelrod-Python,
are covered in Section~\ref{section:data_collection}.
Section~\ref{section:top_performances}, focuses on the best performing
strategies for each type of tournament and overall.
Section~\ref{section:evaluation_of_performance}, explores the traits which
contribute to good performance, and finally the results are summarised in
Section~\ref{section:conclusion}. This manuscripts uses several parameters.
These are introduced in the following sections, however, the full set of
parameters and their definitions are given in Appendix~\ref{app:parameters}.

@@

Figure~\ref{fig:std_results} gives the distributions of $r$ for the top
ranked strategies. The distributions are skewed towards zero and the highest
median, of the top 15 strategies, is at 0.075. This indicates that the top ranked
strategies perform well in any given standard tournament, despite the opponents and
the number of turns.

@@

The top strategies in noisy tournaments are shown in Figure~\ref{fig:noisy_results}. These include deterministic strategies, such
as Tit For 2 Tats~\cite{Axelrod1980b}, Slow Tit For Two Tats~\cite{axelrodproject}, Hard Tit For 2 Tats~\cite{Stewart2012}
and Cycler CCCCCD, and strategies which decide their actions based on the
cooperations to defections ratio, such as ShortMem~\cite{Carvalho2013}, Grumpy
and $e$~\cite{axelrodproject}. Slow Tit For Two Tats is the same strategy as 
Tit For 2 Tats, and at the time of writing this manuscript the
contributors of~\cite{axelrodproject} made a new release where the strategy
has been removed. However, for the purpose of this work the strategy is kept.
The Retaliate and
Limited Retaliate strategies are implemented in APL by the
same contributor. They are strategies designed to defect if the opponent has
tricked them more often than \(x\%\) of the times that they have
done the same. Finally, in $4^{\text{th}}$ and $9^{\text{th}}$ place are Hunter
strategies which trying to extort, equivalently, strategies that play cyclically
and defectors.

@@

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[h]
    \begin{center}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{gcgcgc}
    \toprule
    measure & measure explanation &  source & value type & min value & max value \\
    \midrule
stochastic  &  If a strategy is stochastic & strategy classifier from APL & boolean  & Na &  Na \\
makes use of game &  If a strategy makes used of the game information & strategy classifier from APL & boolean  & Na &  Na \\
makes use of length &  If a strategy makes used of the number of turns & strategy classifier from APL & boolean  & Na &  Na \\
memory usage &  The memory size of a strategy divided by the number of turns & memory size from APL & float & 0 &  1 \\
SSE & A measure of how far a strategy is from ZD behaviour & method described in~\cite{Knight2019} & float & 0 & 1 \\
max cooperating rate $(C_{\text{max}})$  & The biggest cooperating rate in a given tournament  & result summary  & float & 0 & 1\\
min cooperating rate $(C_{\text{min}})$ & The smallest cooperating rate in a given tournament  & result summary  & float & 0 & 1\\
median cooperating rate $(C_{\text{median}})$ & The median cooperating rate in a given tournament  & result summary  & float & 0 & 1\\
mean cooperating rate $(C_{\text{mean}})$ & The mean cooperating rate in a given tournament  & result summary  & float & 0 & 1 \\
$C_r$ / $C_{\text{max}}$ & A strategy's cooperating rate divided by the maximum & result summary  & float & 0 & 1 \\
$C_r$ / $C_{\text{min}}$ & A strategy's cooperating rate divided by the minimum & result summary  & float & 0 & 1 \\
$C_r$ / $C_{\text{median}}$ & A strategy's cooperating rate divided by the median  & result summary  & float & 0 & 1\\
$C_r$ / $C_{\text{mean}}$ & A strategy's cooperating rate divided by the mean & result summary  & float & 0 & 1 \\
$C_r$ & The cooperating ratio of a strategy & result summary  & float & 0 & 1 \\
$CC$ to $C$ rate & The probability a strategy will cooperate after a mutual cooperation & result summary  & float & 0 & 1\\
$CD$ to $C$ rate & The probability a strategy will cooperate after being betrayed by the opponent & result summary  & float & 0 & 1 \\
$DC$ to $C$ rate & The probability a strategy will cooperate after betraying the opponent & result summary  & float & 0 & 1 \\
$DD$ to $C$ rate & The probability a strategy will cooperate after a mutual defection & result summary  & float & 0 & 1 \\
$p_n$ & The probability of a player's action being flip at each interaction & trial summary & float & 0 & 1 \\
$n$ & The number of turns & trial summary & integer & 1 & 200 \\
$p_e$ & The probability of a match ending in the next turn & trial summary & float & 0 & 1 \\
$N$ & The number of strategies in the tournament & trial summary & integer & 3 & 195 \\
$k$ & The number of repetitions of a given tournament & trial summary & integer & 10 & 100 \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{The measures which are included in the performance evaluation analysis.}
    \label{table:manual_measures}
\end{table}

@@

The \numberofstrategies strategies used in this manuscript have been mainly for
the literature, and they have been accessible due to an open source software
called Axelrod-Python library. The software was used to generate a total of
\numberofalltournaments computer tournaments results with different number of
strategies and different participants each time. The data collection was
described in Section~\ref{section:data_collection}. In
Section~\ref{section:top_performances}, the tournaments results were used to
present the top performances. The data set contained results from four different
settings, and these were also studied individually. In standard tournaments
complex strategies trained using reinforcement learning ranked in the top spots.
Some of these strategies ranked again in the top spots in probabilistic
ending tournaments when a \(p_e\) of less 0.1 was considered. In probabilistic
ending tournaments \(p_e\) was designed to vary between 0 and 1. It was demonstrated
that for values larger than 0.1, as stated in the Folk Theorem, defecting strategies
were winning the tournaments because there was a high likelihood of the game
ending in the next turn. In tournaments with noise the median ranks of the top
15 strategies had the highest values and the \(r\) distributions were bimodal.
The top rank strategies were performing both well and bad, and this indicates
that in noisy tournaments there are not strategies that can guarantee winning.
Overall, the top ranked strategies differed from one tournament type to
another and the mechanism behind the winning strategies were all different.
Even strategies designed to do good in one setting did better in others.
On the whole ... (the ipd interactions are unique and there is no winning
strategy) %TODO help from Vince

Section~\ref{section:evaluation_of_performance}, covered an analysis of
performance based on several measures associated with a strategy and with the
environments it was competing. The results of this analysis showed that a
strategy's characteristics such as whether or not it's stochastic, and the information it
used regarding the game had no effect on the strategy's success. The most
important factors have been those that compared the strategy's behaviour to it's
environment. The cooperating ratio of the strategy compared to the mean and
median cooperator was highlighted as the most important feature in the analysis.
More specifically, if a strategy were to enter a tournament with a theory of
mind of its environment it would choose to be the median cooperator in standard
tournaments, to cooperate 10\% in probabilistic ending tournaments and 
60\% in noisy and noisy probabilistic
tournaments of the times the median cooperator did. Lastly, if a strategy was aware of the opponents but not of the
setting on the tournament, a strategy would be more likely to be successful if
it were to identify the median cooperator and cooperated 67\% of the times that
they did.

The data set described in this work contains the largest number of IPD tournaments,
to the authors knowledge, and it available at~\cite{data}. Further data mining
could be applied and provide new insights in the field.

@@

\begin{table}[h]
    \begin{center}
    \resizebox{.7\textwidth}{!}{
    \begin{tabular}{llc}
    \toprule
measure & measure explanation \\
 \midrule
stochastic                  & If a strategy is stochastic \\
makes use of game           & If a strategy makes used of the game information  \\
makes use of length         & If a strategy makes used of the number of turns \\
memory usage                & The memory size of a strategy divided by the number of turns \\
SSE                         & A measure of how far a strategy is from extortionate behaviour \\
$C_{\text{max}}$          & The biggest cooperating rate in the tournament \\
$C_{\text{min}}$          & The smallest cooperating rate in the tournament \\
$C_{\text{median}}$       & The median cooperating rate in the tournament \\
$C_{\text{mean}}$         & The mean cooperating rate in the tournament \\
$C_r$ / $C_{\text{max}}$    & A strategy's cooperating rate divided by the maximum \\
$C_r$ / $C_{\text{min}}$    & A strategy's cooperating rate divided by the minimum \\
$C_r$ / $C_{\text{median}}$ & A strategy's cooperating rate divided by the median \\
$C_r$ / $C_{\text{mean}}$   & A strategy's cooperating rate divided by the mean \\
$C_r$                       & The cooperating ratio of a strategy \\
$CC$ to $C$ rate            & The probability a strategy will cooperate after a mutual cooperation \\
$CD$ to $C$ rate            & The probability a strategy will cooperate after being betrayed by the opponent \\
$DC$ to $C$ rate            & The probability a strategy will cooperate after betraying the opponent\\
$DD$ to $C$ rate            & The probability a strategy will cooperate after a mutual defection\\
$p_n$                       & The probability of a player's action being flip at each interaction \\
$n$                         & The number of turns \\
$p_e$                       & The probability of a match ending in the next turn \\
$N$                         & The number of strategies in the tournament \\
$k$                         & The number that a given tournament is repeated \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{The measures which are included in the performance evaluation analysis.}
\end{table}

>>>>

Though only a select pieces of work have been discussed, the IPD literature is
rich, and new strategies and competitions are being published every year~\cite{Glynatsi2019}. The
question, however, still remains the same: what is the best way to play the
game? Compared to other works, whereas a few selected strategies are evaluated
on a small number of tournaments, this manuscript evaluates the performance of \numberofstrategies
strategies in \numberofalltournaments tournaments. These tournaments do not
consist of just standard round robin tournaments, but also tournaments with noise
and tournaments with a probabilistic ending. The later part of the paper, evaluates
the impact of features on the performance of the strategies using modern
machine learning techniques. These features include measures regarding a
strategy's behaviour and measures regarding the tournaments. The data set used
in this work has been made publicly available~\cite{data} and can be used
for further analysis and insights.

The different tournament types as well as the data collection, which is made
possible due to an open source package called Axelrod-Python,
are covered in Section~\ref{section:data_collection}.
Section~\ref{section:top_performances}, focuses on the best performing
strategies for each type of tournament and overall.
Section~\ref{section:evaluation_of_performance}, explores the traits which
contribute to good performance, and finally the results are summarised in
Section~\ref{section:conclusion}. This manuscripts uses several parameters.
These are introduced in the following sections, however, the full set of
parameters and their definitions are given in Appendix~\ref{app:parameters}.
A list of all strategies including citations of their origins is given in
Appendix~\ref{app:list_of_players}.

@@

Figure~\ref{fig:std_results} gives the distributions of $r$ for the top
ranked strategies. The distributions are skewed towards zero and the highest
median, of the top 15 strategies, is at 0.075. This indicates that the top ranked
strategies perform well in any given standard tournament, regardless of the opponents and
the number of turns.

@@

The top strategies in noisy tournaments are shown in Figure~\ref{fig:noisy_results}. These include deterministic strategies, such
as Tit For 2 Tats~\cite{Axelrod1980b}, Slow Tit For Two Tats~\cite{axelrodproject}, Hard Tit For 2 Tats~\cite{Stewart2012}
and Cycler CCCCCD, and strategies which decide their actions based on the
cooperations to defections ratio, such as ShortMem~\cite{Carvalho2013}, Grumpy
and $e$~\cite{axelrodproject}. Slow Tit For Two Tats is the same strategy as 
Tit For 2 Tats, and at the time of writing this manuscript the
contributors of~\cite{axelrodproject} made a new release where the strategy
has been removed. However, for the purpose of this work the strategy is kept.
The Retaliate and
Limited Retaliate strategies are implemented in APL by the
same contributor. They are strategies designed to defect if the opponent has
tricked them more often than \(x\%\) of the times that they have
done the same. Finally, in $4^{\text{th}}$ and $9^{\text{th}}$ place are Hunter
strategies which try to extort strategies that play cyclically
and defectors.

@@

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[h]
    \begin{center}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{gcgcgc}
    \toprule
    measure & measure explanation &  source & value type & min value & max value \\
    \midrule
stochastic  &  If a strategy is stochastic & strategy classifier from APL & boolean  & Na &  Na \\
makes use of game &  If a strategy makes used of the game information & strategy classifier from APL & boolean  & Na &  Na \\
makes use of length &  If a strategy makes used of the number of turns & strategy classifier from APL & boolean  & Na &  Na \\
memory usage &  The memory size of a strategy divided by the number of turns & memory size from APL & float & 0 &  1 \\
SSE & A measure of how far a strategy is from ZD behaviour & method described in~\cite{Knight2019} & float & 0 & 1 \\
max cooperating rate $(C_{\text{max}})$  & The biggest cooperating rate in a given tournament  & result summary  & float & 0 & 1\\
min cooperating rate $(C_{\text{min}})$ & The smallest cooperating rate in a given tournament  & result summary  & float & 0 & 1\\
median cooperating rate $(C_{\text{median}})$ & The median cooperating rate in a given tournament  & result summary  & float & 0 & 1\\
mean cooperating rate $(C_{\text{mean}})$ & The mean cooperating rate in a given tournament  & result summary  & float & 0 & 1 \\
$C_r$ / $C_{\text{max}}$ & A strategy's cooperating rate divided by the maximum & result summary  & float & 0 & 1 \\
$C_r$ / $C_{\text{min}}$ & A strategy's cooperating rate divided by the minimum & result summary  & float & 0 & 1 \\
$C_r$ / $C_{\text{median}}$ & A strategy's cooperating rate divided by the median  & result summary  & float & 0 & 1\\
$C_r$ / $C_{\text{mean}}$ & A strategy's cooperating rate divided by the mean & result summary  & float & 0 & 1 \\
$C_r$ & The cooperating ratio of a strategy & result summary  & float & 0 & 1 \\
$CC$ to $C$ rate & The probability a strategy will cooperate after a mutual cooperation & result summary  & float & 0 & 1\\
$CD$ to $C$ rate & The probability a strategy will cooperate after being betrayed by the opponent & result summary  & float & 0 & 1 \\
$DC$ to $C$ rate & The probability a strategy will cooperate after betraying the opponent & result summary  & float & 0 & 1 \\
$DD$ to $C$ rate & The probability a strategy will cooperate after a mutual defection & result summary  & float & 0 & 1 \\
$p_n$ & The probability of a player's action being flip at each interaction & trial summary & float & 0 & 1 \\
$n$ & The number of turns & trial summary & integer & 1 & 200 \\
$p_e$ & The probability of a match ending in the next turn & trial summary & float & 0 & 1 \\
$N$ & The number of strategies in the tournament & trial summary & integer & 3 & 195 \\
$k$ & The number of repetitions of a given tournament & trial summary & integer & 10 & 100 \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{The features which are included in the performance evaluation analysis.}
    \label{table:manual_measures}
\end{table}

@@

The \numberofstrategies strategies used in this manuscript have been mainly from
the literature, and they have been accessible due to an open source software
called Axelrod-Python library. The software was used to generate a total of
\numberofalltournaments computer tournaments results with different number of
strategies and different participants each time. The data collection was
described in Section~\ref{section:data_collection}. In
Section~\ref{section:top_performances}, the tournaments results were used to
present the top performances. The data set contained results from four different
settings, and these were also studied individually. In standard tournaments
complex strategies trained using reinforcement learning ranked in the top spots.
Some of these strategies ranked again in the top spots in probabilistic
ending tournaments when a \(p_e\) of less 0.1 was considered. In probabilistic
ending tournaments \(p_e\) was designed to vary between 0 and 1. It was demonstrated
that for values larger than 0.1, as stated in the Folk Theorem, defecting strategies
were winning the tournaments because there was a high likelihood of the game
ending in the next turn. In tournaments with noise the median ranks of the top
15 strategies had the highest values and the \(r\) distributions were bimodal.
The top rank strategies were performing both well and bad, and this indicates
that in noisy tournaments there are not strategies that can guarantee winning.
Overall, the top ranked strategies differed from one tournament type to
another and the mechanism behind the winning strategies were all different.
Even strategies designed to do good in one setting did better in others.
On the whole ... (the ipd interactions are unique and there is no winning
strategy) %TODO help from Vince

Section~\ref{section:evaluation_of_performance}, covered an analysis of
performance based on several measures associated with a strategy and with the
environments it was competing. The results of this analysis showed that a
strategy's characteristics such as whether or not it's stochastic, and the information it
used regarding the game had no effect on the strategy's success. The most
important factors have been those that compared the strategy's behaviour to it's
environment. The cooperating ratio of the strategy compared to the mean and
median cooperator was highlighted as the most important feature in the analysis.
More specifically, if a strategy were to enter a tournament with a theory of
mind of its environment it would choose to be the median cooperator in standard
tournaments, to cooperate 10\% of the times the median cooperator did in probabilistic ending tournaments and 
60\% in noisy and noisy probabilistic
tournaments. Lastly, if a strategy was aware of the opponents but not of the
setting on the tournament, a strategy would be more likely to be successful if
it were to identify the median cooperator and cooperated 67\% of the times that
they did.

The data set described in this work contains the largest number of IPD tournaments,
to the authors knowledge, and available at~\cite{data}. Further data mining
could be applied and provide new insights in the field.

@@

A summary of the parameters and features used in this work are given
by Table~\ref{table:parameters_summary}.

\begin{table}[h]
    \begin{center}
    \resizebox{.7\textwidth}{!}{
    \begin{tabular}{llc}
    \toprule
measure & measure explanation \\
 \midrule
stochastic                  & If a strategy is stochastic \\
makes use of game           & If a strategy makes used of the game information  \\
makes use of length         & If a strategy makes used of the number of turns \\
memory usage                & The memory size of a strategy divided by the number of turns \\
SSE                         & A measure of how far a strategy is from extortionate behaviour \\
$C_{\text{max}}$          & The biggest cooperating rate in the tournament \\
$C_{\text{min}}$          & The smallest cooperating rate in the tournament \\
$C_{\text{median}}$       & The median cooperating rate in the tournament \\
$C_{\text{mean}}$         & The mean cooperating rate in the tournament \\
$C_r$ / $C_{\text{max}}$    & A strategy's cooperating rate divided by the maximum \\
$C_r$ / $C_{\text{min}}$    & A strategy's cooperating rate divided by the minimum \\
$C_r$ / $C_{\text{median}}$ & A strategy's cooperating rate divided by the median \\
$C_r$ / $C_{\text{mean}}$   & A strategy's cooperating rate divided by the mean \\
$C_r$                       & The cooperating ratio of a strategy \\
$CC$ to $C$ rate            & The probability a strategy will cooperate after a mutual cooperation \\
$CD$ to $C$ rate            & The probability a strategy will cooperate after being betrayed by the opponent \\
$DC$ to $C$ rate            & The probability a strategy will cooperate after betraying the opponent\\
$DD$ to $C$ rate            & The probability a strategy will cooperate after a mutual defection\\
$p_n$                       & The probability of a player's action being flip at each interaction \\
$n$                         & The number of turns \\
$p_e$                       & The probability of a match ending in the next turn \\
$N$                         & The number of strategies in the tournament \\
$k$                         & The number that a given tournament is repeated \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{The measures which are included in the performance evaluation analysis.}\label{table:parameters_summary}
\end{table}
